---
title: "transform_solutions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(nycflights13)
library(ggplot2)
```

# 4.2.4
### Find all flights that
#### Had an arrival delay of two or more hours.
```{r}
filter(flights, arr_delay>120)
```

#### Flew to Houston (`IAH` or `HOU`)
```{r}
filter(flights, dest %in% c('IAH', 'HOU'))
```

#### Were operated by United, American, or Delta
```{r}
filter(flights, carrier %in% c('UA', 'AA', 'DL'))
```

#### Departed in summer (July, August, and September)
```{r}
filter(flights, month <= 7, month <= 9)
```

#### Arrived more than two hours late, but didn't leave late
```{r}
filter(flights, arr_delay > 120, dep_delay <= 0)
```

#### Were delayed by at least an hour, but made up over 30 minutes in flight
```{r}
filter(flights, dep_delay >= 60, dep_delay-arr_delay > 30)
```

#### Departed between midnight and 6am (inclusive)
```{r}
filter(flights, dep_time <=600 | dep_time == 2400)
```

### Another useful dplyr filtering helper is `between()`. What does it do? Can you use it to simplify the code needed to answer the previous challenges?
Between is a shorter, faster way of testing two inequalities at once: it tests if its first argument is greater than or equal to its second, and less than or equal to its third.
```{r}
filter(flights, between(month, 7, 9))
filter(flights, !between(dep_time, 601, 2359))
```

### How many flights have a missing `dep_time`? What other variables are missing? What might these rows represent?
```{r}
summary(flights)
```

8255 flights have a missing `dep_time`, 8255 have a missing `dep_delay`, 8713 have a missing `arr_time`, 9430 have a missing `arr_delay`, and 9430 have a missing `air_time`. We can speculate that these are flights that failed to depart or arrive, since a flight that departs normally but is then rerouted will probably have a normally recorded departure but no similar record for it's arrival. However, these could also just be lost data about perfectly normal flights.

### Why is `NA ^ 0` not missing? Why is `NA | TRUE` not missing? Why is `FALSE & NA` not missing? Can you figure out the general rule?  (`NA * 0` is a tricky counterexample!)
`NA ^ 0` is 1 because anything to the power of 0 is 1, so though we didn't know the original value, we know it's zeroth power. Similarly with `FALSE & NA`. `NA * 0` could be argued to be because the NA could represent `Inf`, and `Inf * 0` is `NaN`, rather than `NA`. However, I suspect that these results are dictated as much by what answer is natural and quick and sensible in C as by mathematical corner cases.

